Chatbot based Cognitive Behavioural Therapy

### 2021-05-19 paper.tex
minor changes, recommendations

\section{Introduction}
During the last two decades the internet has become a major part in the life
of people throughout the society. The ubiquity has led to the desire for 
location-independent usage. As a consequence the amount of
smartphone users is increasing every year.\footnote{https://www.statista.com/statistics/330695/number-of-smartphone-users-worldwide/, last access: 13.05.2021}
With constant access to the biggest collection of knowledge in the world,
more and more individuals tend to use it as a first source of information when
symptoms of illness arise\cite{Wyatt2015}. 
In terms of mental health, the internet provides a numerous amount of cognitive behavioural therapy (CBT) offers such as guides
or apps. 
\\\\
In contrast to a complex psychoanalysis with the guidance of a therapist or doctor trying to look into the past,
CBT focuses more on the way an individual deals with its present thoughts and beliefs. The treatment has the aim to
change behavioural patterns in a way that they positively influence emotions and cognition. CBT is based on this connection
between the way how we interpret the daily life, how we behave due to it and how we 
feel in the end which is shown in figure 1. By changing one of the three aspects,
the others are changed as well automatically. Some popular methods are keeping a Gratitude Journal or exercises to calm the body.\cite{Spangler2002}
\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{cbt_picture}
  \caption{The relationships among cognition, emotion, behavior. Graphic by Raja Selvam, 2018
    \\(\url{https://integralsomaticpsychology.com/wp-content/uploads/2018/02/ISP-Embodied-Cognition-Diagram-Raja-Selvam-750.jpg}).}
  \Description{The relationships among cognition, emotion, behavior.}
\end{figure}
\\
With the help of apps, users can get hints and instructions to conduct a self therapy. It is possible
to get first support without the need to consult a therapist or doctor. Often, this a big and 
difficult step for many people due to the fear of getting stigmatized\cite{Rossler2016}.
While CBT applications have the potential to reduce the barrier to seek help, the rate of completion
is relatively low. Donkin et. al found a median minimal completion rate of 56\% during a randomized controlled trial\cite{Donkin}. 
One possible reason for this lack of adherence is the missing human interaction of internet-based CBT\cite{Ly2017}.
The usage of chatbots are considered as a possible solution for this problem. While they are providing CBT on the same level 
as non-chatbot based approaches, they can deliver an artificial human interaction that is supposed to increase the
therapy adherence which can be seen as an important aspect of a successful therapy. 
\\\\
The paper starts with a review of a part of the conducted research and analyses weaknesses and strengths as well as risks and chances. 
Subsequently, it will define some general requirements a chatbot used for CBT should met.
What are aspects that need improvement during further development approaches and what methods and measures might be suitable to achive that? 
Considering the state of the art and the general requirements, the paper tries to answer this question regarding safety aspects, efficiacy and privacy.
It finally discusses the outcomes and points out limits of chatbot based CBT.

@@

This far, there are a couple of chatbots available. This paper will mainly focus on the Woebot Chatbot and Wysa for several reasons.

@@

\subsection{Woebot}
The Woebot chatbot is a product of the company Woebot Health which has its headquarter in California. It is the reusult of a resarch 

@@

Woebot is available as a standalone app in Apple's App Store and in Google Play store where it reached more than 100.000 downloads\footnote{https://play.google.com/store/apps/details?id=com.woebot\&hl=de\&gl=US, last access: 03.05.2021} and as 

@@

In general, it gives hints how to change the thinking and guides the user through techniques like gratitude journaling. 

@@

Woebot aims to reach patients and clinicians. While users that are currently not in therapy have the opportunity to use it,

@@

Currently, the usage of woebot is completely free of charge. Nevertheless, Woebot Health announces on its website, that they are moving forward

@@

A randomized controlled trial with college students by Fitzpatrick et. al has shown that the usage of Woebot significantly reduces

@@

After the period of two weeks the users have been asked to assess their experience with Woebot. One the one hand, the majority appreciated the daily check-ins of the bot 

@@

The privacy policy of Woebot gives some useful information about the way the chatbot deals with user inputs. 

@@

Additionally, neither access to stored data nor backup functionalities are possible. If a user creates an account, Woebot Health stores

@@

to improve the services. Woebot Health also obtains data from other sources. If the service is used trough third-party applications, 

@@

aspects. If a user starts a conversation with Woebot, the bot early advises that it is not a human being. It notes that it has limited capabilities and that it might 
be better for some individuals to see a real therapist. In order to discourage over-reliance, Woebot tries to end conversations after a certain amount of time and proposes many non-digital activities in the real world.

@@

Considering emergency situations, Woebot is able to recognize user inputs like 'SOS' or 'suicide' as such. It reacts with a list of helplines

@@

\subsection{Wysa}
The Wysa Chatbot, developed by Touchkin, has certain commonalities but also major differences in comparison to Woebot. It is also a mobile chatbot app using a text-based interface with the possibility of entering 

@@

a human therapist while using Wysa. It is a chargeable feature which costs 29.99\$ per month. The usage of the chatbot without it is free of charge. It is available

@@

While also offering CBT like Woebot, Wysa also offers guided meditation, breathing and yoga. Additionally, it conducts Dialectial Behaviour Therapy (DBT).\textsuperscript{6} 

@@

Inkster et. al have shown that patients with a more frequent use of Wysa have a significantly higher improvment of their mood than users

@@

While the app was generally considered as helpful and encouraging, the participants proposed improvements included wanting to avoid repetitions or a better understanding of user inputs.\cite{Inkster} 

@@

The way Wysa stores user data deviates from Woebot's. Though Wysa does not require a registration, it links all the collected data to a vendor-specific user id provided by Google Play or App store. This approach leads to a pseudonymised data storage.
If a user accidentally submits personal information during a conversation, it will be redacted by algorithms scanning the stored data regularly. In addition to conversational data, Wysa stores anonymized technical data (platform, operating system), 

@@

It is important to mention, that the chatbot can not be used through third party applications like Facebooks's messenger. Hence a user can use the application more independently of privacy policies of other companies. Altough it is not completely excluded as Wysa allows to speak with the chatbot 

@@

Considering privacy aspects, Wysa is quite common to Woebot. It also informs the user during the conversation, that it is only a robot. If an individual is willing to pay the fee, the contact to a human therapist is a safe way when it has mental issues that are 
more serious. But as this service is not free of charge, it can not be seen as a general safety aspect. The Wysa chatbot tries to deal with emergency situations the same way as Woebot does. It notices user inputs like 'SOS' and provides helplines for further support.\cite{Kretzschmar2019}

@@

This chapter will define requirements a chatbot should meet regarding privacy, efficiacy and safety. Those three aspects have already been examined by Kretzschmar et.al, but only from the perspection of young adults as a result of a group discussion\cite{Kretzschmar2019}.

@@

Firstly, it is necessary to define what privacy principles are concerning HCI. In this paper, the results of Patrick and Kenny are used, who specifically derived privacy aspects by using the general EU legislation as basis.

@@

awareness of the individual and that it is informed about data processing constantly. \emph{Control} contains principles to have access to stored data and to be empowered to delete it and requirements of the \emph{consent} category make sure that the user always has to give its approval about processing of personal data.\cite{Patrick2003}

@@

As chatbots are only a subset of HCI and the categories and principles above are held general, it is important to concretise them regarding mental health chatbots. Saglam et. al conducted an empirical study where on the one hand, the participants have been asked to answer questions about their trust towards a chatbot and 

@@

For the second question, the study has revealed that users prefer chatbots with good technical quality like grammatically correct answers rather than those with social aspects like seeing an avatar while chatting.\cite{Shamim2021}

@@

Chatbots with a better response quality can be seen as more human-like. Ischen et. al revealed in another study, that human-like chatbots lead to less privacy concerns\cite{Ischen}.

@@

@@

Another aspect that has not been considered yet is the distinction between data and information. An interview with employees about their attitude towards a workplace robot conducted by Lee et. al showed that most participants did not make a distinction between the knowledge of the robot and the data it collects.\cite{Lee2011}

@@

@@

its algorithms better. Only with all of these aspects given, the individual is able to give its consent on a proper basis. 

@@

Ludden et.al proposed two different routes that can lead to a better well-being that are shown in figure 3. They introduced the PERMA abbrevation which stands for positive emotions, engagement, positive relationships, meaning and accomplishment. All of these aspects can lead to a higher adherence and hence to a better well-being.\cite{Ludden2015}

@@

  \caption{Schematic representation of how the design of a Web-based intervention can influence general subjective well-being following two different routes. Route 1 indicates the impact of design on adherence and thus on the focal health problem. Route 2 indicates how overall well-being is stimulated by elements of PERMA. Graphic by Ludden et.al, 2015}

@@

@@

Donkin et. al found a minimal completion rate of 56\% concerning online-therapy applications. Furthermore they revealed that the adherence correlates with the treatment outcomes.\cite{Donkina}
\\This can be considered as a further prove for Route 1 in figure 3.  

@@

Ludden et. al present personalization as one approach. Considering the PERMA abbrevation, personalization encourage the engagement of a user as the application has a higher personal relevance. If it is used by multiple groups of people, it should be adaptable to them in order to improve the experience.

@@

They note that the way how human beings think about abstract things like challenges or achivements is methaphorical by nature and experienced every day. Metaphors supporting mental health applications can strengthen those experiences.\cite{Ludden2015}

@@

According to Ischen et.al, services with a higher perception in anthropomorphism lead to a better adherence \cite{Ischen}.

@@

Kretzschmar et. al defined four requirements that are essential. Firstly, it is considered as crucial that users get informed about the fact that they are talking to a robot. This makes sure that they are aware of limitations and that the conversation can not replace a real life therapy.

@@

Kretzschmar et. al finally value the recognition of emergency situations.\cite{Kretzschmar2019}

@@

This last aspect has also been recommendated by C.Grové who co-developed a chatbot with and for young people. She points out that the identification of critical words is necessary for a chatbot that deals with mental health.

@@

As a second aspect she notes that the applications should have a sufficient listening phase prior to quick responses that might not be suitable to the situation.\cite{Grove2021}

>>>>

\section{Introduction}
During the last two decades the internet has become a major part in the life
of people throughout the society. The ubiquity has led to the desire for 
location-independent usage. As a consequence the amount of
smartphone users is increasing every year.\footnote{https://www.statista.com/statistics/330695/number-of-smartphone-users-worldwide/, last access: 13.05.2021}
With constant access to the biggest collection of knowledge in the world,
more and more individuals tend to use it as a first source of information when
symptoms of illness arise\cite{Wyatt2015}. 
In terms of mental health, the internet provides a numerous amount of \emph{cognitive behavioural therapy} (CBT) offers such as guides
or apps. 
\\\\
In contrast to a complex psychoanalysis with the guidance of a therapist or doctor trying to look into the past,
CBT focuses more on the way an individual deals with its present thoughts and beliefs. The treatment has the aim to
change behavioural patterns in a way that they positively influence emotions and cognition. CBT is based on this connection
between the way how we interpret the daily life, how we behave due to it and how we 
feel in the end which is shown in Figure \ref{fig:cbt_circle}. By changing one of the three aspects,
the others are changed as well automatically. Some popular methods are keeping a \emph{Gratitude Journal} or exercises to calm the body.\cite{Spangler2002}
\\
\emph{Gratitude Journaling} works by writing down moments, situations or facts the writer is grateful for. It can lead to a more positive mindset and higher mindfulness. 
\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{cbt_picture}
  \caption[cbt_circle]{The relationships among cognition, emotion, behavior. Graphic by Raja Selvam, 2018\footnotemark}
  \Description{The relationships among cognition, emotion, behavior.}
  \label{fig:cbt_circle}
\end{figure}
\footnotetext{https://integralsomaticpsychology.com/wp-content/uploads/2018/02/ISP-Embodied-Cognition-Diagram-Raja-Selvam-750.jpg, last access: 30.04.2021}
\\
With the help of apps, users can get hints and instructions to conduct self-help. It is possible
to get first support without the need to consult a therapist or doctor. Often, this a big and 
difficult step for many people due to the fear of getting stigmatized\cite{Rossler2016}.
While CBT applications have the potential to reduce the barrier to seek help, the rate of completion
is relatively low. Donkin et. al \cite{Donkin} found a median minimal completion rate of 56\% during a randomized controlled trial. 
One possible reason for this lack of adherence is the missing human interaction of internet-based CBT\cite{Ly2017}.
The usage of chatbots are considered as a possible solution for this problem. While they are providing CBT on the same level 
as non-chatbot based approaches, they can deliver an artificial human interaction that is supposed to increase the
therapy adherence which can be seen as an important aspect of a successful therapy. 
\\\\
The paper starts with a review of a part of the conducted research and analyses weaknesses and strengths as well as risks and chances. 
Subsequently, it will define some general requirements a chatbot used for CBT should met.
What are aspects that need improvement during further development approaches and what methods and measures might be suitable to achive that? 
Considering the state of the art and the general requirements, the paper tries to answer this question regarding safety aspects, efficiacy and privacy.
It finally discusses the outcomes and points out limits of chatbot based CBT.

@@

This far, there are a couple of chatbots available. This paper will mainly focus on the \emph{Woebot} Chatbot and \emph{Wysa} for several reasons.

@@

\subsection{\emph{Woebot}}
The \emph{Woebot} chatbot is a product of the company \emph{WoebotHealth} which has its headquarter in California. It is the reusult of a resarch 

@@

\emph{Woebot} is available as a standalone app in Apple's App Store and in Google Play store where it reached more than 100.000 downloads\footnote{https://play.google.com/store/apps/details?id=com.woebot\&hl=de\&gl=US, last access: 03.05.2021} and as 

@@

In general, it gives hints how to change the thinking and guides the user through techniques like \emph{Gratitude Journaling}. 

@@

\emph{Woebot} aims to reach patients and clinicians. While users that are currently not in therapy have the opportunity to use it,

@@

Currently, the usage of \emph{Woebot} is completely free of charge. Nevertheless, \emph{Woebot} Health announces on its website, that they are moving forward

@@

A randomized controlled trial with college students by Fitzpatrick et. al \cite{Fitzpatrick2017} has shown that the usage of \emph{Woebot} significantly reduces

@@

After the period of two weeks the users have been asked to assess their experience with \emph{Woebot}. One the one hand, the majority appreciated the daily check-ins of the bot 

@@

The privacy policy of \emph{Woebot} gives some useful information about the way the chatbot deals with user inputs. 

@@

Additionally, neither access to stored data nor backup functionalities are possible. If a user creates an account, \emph{WoebotHealth} stores

@@

to improve the services. \emph{WoebotHealth} also obtains data from other sources. If the service is used trough third-party applications, 

@@

aspects. If a user starts a conversation with \emph{Woebot}, the bot early advises that it is not a human being. It notes that it has limited capabilities and that it might 
be better for some individuals to see a real therapist. In order to discourage over-reliance, \emph{Woebot} tries to end conversations after a certain amount of time and proposes many non-digital activities in the real world.

@@

Considering emergency situations, \emph{Woebot} is able to recognize user inputs like 'SOS' or 'suicide' as such. It reacts with a list of helplines

@@

\subsection{\emph{Wysa}}
The \emph{Wysa} Chatbot, developed by \emph{Touchkin}, has certain commonalities but also major differences in comparison to \emph{Woebot}. It is also a mobile chatbot app using a text-based interface with the possibility of entering 

@@

a human therapist while using \emph{Wysa}. It is a chargeable feature which costs 29.99\$ per month. The usage of the chatbot without it is free of charge. It is available

@@

While also offering CBT like \emph{Woebot}, \emph{Wysa} also offers guided meditation, breathing and yoga. Additionally, it conducts \emph{Dialectial Behaviour Therapy} (DBT).\textsuperscript{6} 

@@

Inkster et. al \cite{Inkster} have shown that patients with a more frequent use of \emph{Wysa} have a significantly higher improvment of their mood than users

@@

While the app was generally considered as helpful and encouraging, the participants proposed improvements included wanting to avoid repetitions or a better understanding of user inputs.

@@

The way \emph{Wysa} stores user data deviates from \emph{Woebot}'s. Though \emph{Wysa} does not require a registration, it links all the collected data to a vendor-specific user id provided by Google Play or App store. This approach leads to a pseudonymised data storage.
If a user accidentally submits personal information during a conversation, it will be redacted by algorithms scanning the stored data regularly. In addition to conversational data, \emph{Wysa} stores anonymized technical data (platform, operating system), 

@@

It is important to mention, that the chatbot can not be used through third party applications like Facebooks's messenger. Hence a user can use the application more independently of privacy policies of other companies. Altough it is not completely excluded as \emph{Wysa} allows to speak with the chatbot 

@@

Considering privacy aspects, \emph{Wysa} is quite common to \emph{Woebot}. It also informs the user during the conversation, that it is only a robot. If an individual is willing to pay the fee, the contact to a human therapist is a safe way when it has mental issues that are 
more serious. But as this service is not free of charge, it can not be seen as a general safety aspect. The \emph{Wysa} chatbot tries to deal with emergency situations the same way as \emph{Woebot} does. It notices user inputs like 'SOS' and provides helplines for further support.\cite{Kretzschmar2019}

@@

This chapter will define requirements a chatbot should meet regarding privacy, efficiacy and safety. Those three aspects have already been examined by Kretzschmar et.al \cite{Kretzschmar2019}, but only from the perspection of young adults as a result of a group discussion.

@@

Firstly, it is necessary to define what privacy principles are concerning HCI. In this paper, the results of Patrick and Kenny \cite{Patrick2003} are used, who specifically derived privacy aspects by using the general EU legislation as basis.

@@

awareness of the individual and that it is informed about data processing constantly. \emph{Control} contains principles to have access to stored data and to be empowered to delete it and requirements of the \emph{consent} category make sure that the user always has to give its approval about processing of personal data.

@@

As chatbots are only a subset of HCI and the categories and principles above are held general, it is important to concretise them regarding mental health chatbots.
Saglam et. al \cite{Shamim2021} conducted an empirical study where on the one hand, the participants have been asked to answer questions about their trust towards a chatbot and 

@@

For the second question, the study has revealed that users prefer chatbots with good technical quality like grammatically correct answers rather than those with social aspects like seeing an avatar while chatting.

@@

Chatbots with a better response quality can be seen as more human-like. Ischen et. al \cite{Ischen} revealed in another study, that human-like chatbots lead to less privacy concerns.

@@

Another aspect that has not been considered yet is the distinction between data and information.

@@

An interview with employees about their attitude towards a workplace robot conducted by Lee et. al \cite{Lee2011} showed that most participants did not make a distinction between the knowledge of the robot and the data it collects.

@@

  \label{fig:privacy_base}

@@

its algorithms better. Only with all of these aspects given, the individual is able to give its consent on a proper basis as shown in Figure \ref{fig:privacy_base}. 

@@

Ludden et.al \cite{Ludden2015} proposed two different routes that can lead to a better well-being that are shown in Figure \ref{fig:routes}. They introduced the PERMA abbrevation which stands for positive emotions, engagement, positive relationships, meaning and accomplishment. All of these aspects can lead to a higher adherence and hence to a better well-being.

@@

  \caption{Schematic representation of how the design of a Web-based intervention can influence general subjective well-being following two different routes. Route 1 indicates the impact of design on adherence and thus on the focal health problem. Route 2 indicates how overall well-being is stimulated by elements of PERMA. Graphic by Ludden et.al \cite{Ludden2015}, 2015}

@@

  \label{fig:routes}

@@

Donkin et. al \cite{Donkina} found a minimal completion rate of 56\% concerning online-therapy applications. Furthermore they revealed that the adherence correlates with the treatment outcomes.
\\This can be considered as a further prove for Route 1 in Figure \ref{fig:routes}.  

@@

Ludden et. al \cite{Ludden2015} present personalization as one approach. Considering the PERMA abbrevation, personalization encourage the engagement of a user as the application has a higher personal relevance. If it is used by multiple groups of people, it should be adaptable to them in order to improve the experience.

@@

They note that the way how human beings think about abstract things like challenges or achivements is methaphorical by nature and experienced every day. Metaphors supporting mental health applications can strengthen those experiences.

@@

According to Ischen et.al \cite{Ischen}, services with a higher perception in anthropomorphism lead to a better adherence.

@@

Kretzschmar et. al \cite{Kretzschmar2019} defined four requirements that are essential. Firstly, it is considered as crucial that users get informed about the fact that they are talking to a robot. This makes sure that they are aware of limitations and that the conversation can not replace a real life therapy.

@@

Kretzschmar et. al finally value the recognition of emergency situations.

@@

This last aspect has also been recommended by Grové \cite{Grove2021} who co-developed a chatbot with and for young people. She points out that the identification of critical words is necessary for a chatbot that deals with mental health.

@@

As a second aspect she notes that the applications should have a sufficient listening phase prior to quick responses that might not be suitable to the situation.
